---
title: "Filtering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(stringr)
```

```{r clear environment and import data}

rm(list = ls()) 

# define data directory
data_dir <- "path_to_folder/"

# load data (IFNG, but the same cells are also in TNF) with cell and nucleus area for different timepoints 
Area_0H <- read.csv(paste0(data_dir, "0H/IFN__FQ_batch_summary_MATURE_221205.txt"), header=F, sep="", skip = 5, col.names = c("file", "cell", "AREA_cell", "AREA_nucleus", "N_total",	"N_thres_Total",	"N_thres_Nuc"))
Area_1H <- read.csv(paste0(data_dir, "1H/IFN__FQ_batch_summary_MATURE_221130.txt"), header=F, sep="", skip = 5, col.names = c("file", "cell", "AREA_cell", "AREA_nucleus", "N_total",	"N_thres_Total",	"N_thres_Nuc"))
Area_2H <- read.csv(paste0(data_dir, "2H/IFN__FQ_batch_summary_MATURE_221207.txt"), header=F, sep="", skip = 5, col.names = c("file", "cell", "AREA_cell", "AREA_nucleus", "N_total",	"N_thres_Total",	"N_thres_Nuc"))
Area_3H <- read.csv(paste0(data_dir, "3H/IFN__FQ_batch_summary_MATURE_221207.txt"), header=F, sep="", skip = 5, col.names = c("file", "cell", "AREA_cell", "AREA_nucleus", "N_total",	"N_thres_Total",	"N_thres_Nuc"))
Area_4H <- read.csv(paste0(data_dir, "4H/IFN__FQ_batch_summary_MATURE_221208.txt"), header=F, sep="", skip = 5, col.names = c("file", "cell", "AREA_cell", "AREA_nucleus", "N_total",	"N_thres_Total",	"N_thres_Nuc"))
Area_6H <- read.csv(paste0(data_dir, "6H/IFN__FQ_batch_summary_MATURE_221208.txt"), header=F, sep="", skip = 5, col.names = c("file", "cell", "AREA_cell", "AREA_nucleus", "N_total",	"N_thres_Total",	"N_thres_Nuc"))


# fetch all files in FQ_outline folder
Outlines_0H = list.files(paste0(data_dir, "0H/FQ_outline"), full.names=TRUE)
Outlines_1H = list.files(paste0(data_dir, "1H/FQ_outline"), full.names=TRUE)
Outlines_2H = list.files(paste0(data_dir, "2H/FQ_outline"), full.names=TRUE)
Outlines_3H = list.files(paste0(data_dir, "3H/FQ_outline"), full.names=TRUE)
Outlines_4H = list.files(paste0(data_dir, "4H/FQ_outline"), full.names=TRUE)
Outlines_6H = list.files(paste0(data_dir, "6H/FQ_outline"), full.names=TRUE)

# identify and fetch the log file among the files by matching "Logfile" with file names; make it case insensitive 
Logfile_0H = Outlines_0H[which(grepl("Logfile", Outlines_0H, ignore.case = T))]
Logfile_1H = Outlines_1H[which(grepl("Logfile", Outlines_1H, ignore.case = T))]
Logfile_2H = Outlines_2H[which(grepl("Logfile", Outlines_2H, ignore.case = T))]
Logfile_3H = Outlines_3H[which(grepl("Logfile", Outlines_3H, ignore.case = T))]
Logfile_4H = Outlines_4H[which(grepl("Logfile", Outlines_4H, ignore.case = T))]
Logfile_6H = Outlines_6H[which(grepl("Logfile", Outlines_6H, ignore.case = T))]

# remove logfile from the bulk of files; now outline has only outline files
Outlines_0H = setdiff(Outlines_0H, Logfile_0H)
Outlines_1H = setdiff(Outlines_1H, Logfile_1H)
Outlines_2H = setdiff(Outlines_2H, Logfile_2H)
Outlines_3H = setdiff(Outlines_3H, Logfile_3H)
Outlines_4H = setdiff(Outlines_4H, Logfile_4H)
Outlines_6H = setdiff(Outlines_6H, Logfile_6H)


# fetch quantification files
Spots_0H <- list.files(paste0(data_dir, "0H/3D_spots"), full.names=TRUE)
Spots_1H <- list.files(paste0(data_dir, "1H/3D_spots"), full.names=TRUE)
Spots_2H <- list.files(paste0(data_dir, "2H/3D_spots"), full.names=TRUE)
Spots_3H <- list.files(paste0(data_dir, "3H/3D_spots"), full.names=TRUE)
Spots_4H <- list.files(paste0(data_dir, "4H/3D_spots"), full.names=TRUE)
Spots_6H <- list.files(paste0(data_dir, "6H/3D_spots"), full.names=TRUE)


rm(data_dir)

```

```{r identify cells with a nucleus}

# combine all timepoints into 1 dataframe and remove all rows with nuclear are == 0
Area_all = rbind(Area_0H, Area_1H, Area_2H, Area_3H, Area_4H, Area_6H)
Area_all <- Area_all[Area_all$AREA_nucleus>0 ,]

# make ID that combines all important info to uniquely determine a cell
Area_all <- Area_all %>% dplyr::mutate(ID = substr(Area_all$file, 1, 15))
Area_all <- Area_all %>% dplyr::mutate(ID = paste0(ID, "_", cell))

# rename dataframe and keep only ID column; remove all data frames that we don't need anymore
Cells_with_nucleus <- Area_all %>% select(ID)
rm(Area_all, Area_0H, Area_1H, Area_2H, Area_3H, Area_4H, Area_6H)

```

IDs of cells with a nucleus (from all timepoints) are in dataframe Cells_with_nucleus

```{r identify cells in division}

# use logfile from fish quant to identify cells that are in division

# combine all logfiles into a list
logfiles = c(Logfile_0H, Logfile_1H, Logfile_2H, Logfile_3H, Logfile_4H, Logfile_6H)

ID = ""
Dividing_cells = data.frame()

# go one by one logfile
for (file_ in logfiles) {
  
  # open logfile 
  log = read.csv(file_, sep = '\t', header = FALSE)
  
  # go line by line of logfile
  for (i in 1:dim(log)[1]){
  
  # if line has ANALYZE file and two lines below is RUNNING conversion, then 1 line below extract file name (ID)
  if (grepl("=== ANALYZE file", log[i,1]) == TRUE){
    if (grepl("RUNNING CONVERSION for cells ...", log[i+2,1]) == TRUE){
    ID = substr(log[i+1,1], 7,21)
    } 
  }
  
   # if line contains Cell already has a nucleus, take cell ID from one line below, and together with ID put into data frame
   if (grepl("Cell already has a nucleus", log[i,1]) == TRUE){
     Dividing_cells <- rbind(Dividing_cells, c(ID, substr(log[i+1,1],12,14)))
   }
  
  }
}

rm(log, logfiles, Logfile_0H, Logfile_1H, Logfile_2H, Logfile_3H, Logfile_3H, Logfile_4H, Logfile_6H, i, ID, file_)

# rename column names, combine file and cell info, and then keep only this unique ID
colnames(Dividing_cells) <- c("file", "cell")
Dividing_cells <- Dividing_cells %>% dplyr::mutate(ID = paste0(file, "_Cell_CP_", cell))
Dividing_cells <- Dividing_cells %>% select(ID)

# remove duplicated rows - could be that a cell had more than 2 nuclei so appears more than once here
Dividing_cells <- as.data.frame(Dividing_cells[!duplicated(Dividing_cells) ,])
colnames(Dividing_cells) <- "ID"

write.csv(Dividing_cells, "List_of_DIVIDING_CELLS.txt")

```

IDs of cells that are in division (from all timepoints) are in dataframe Dividing_cells

```{r identify cells that are on edges of coverslips}

cell = ""
Cells_on_edge = data.frame()

Outlines_all <- c(Outlines_0H, Outlines_1H, Outlines_2H, Outlines_3H, Outlines_4H, Outlines_6H)

for (file_ in Outlines_all) {

  #print(file_)
  outline = read.csv(file_, sep = ' ', header = FALSE)

  for (i in 1:dim(outline)[1]){
  
    # name of file
    if (grepl("IMG_Raw", outline[i,1]) == TRUE){
      file = substr(outline[i,1], 9, 23)
    }
    
     # if cell coordinates are starting, note the cell ID and look for edge coordinates
     if (grepl("CELL_START", outline[i,1]) == TRUE){
       cell = substr(outline[i,1], 12, 22)
       edge = "no"
       
       #check if in X_POS or Y_POS (one and two rows below) there's something touching the edge
       if (grepl("\t1\t|\t2304", outline[i+1,1]) == TRUE | grepl("\t1\t|\t2304", outline[i+2,1]) == TRUE){
         edge = "yes"
       }
       
       if (edge == "yes"){
         Cells_on_edge <- rbind(Cells_on_edge, c(file, cell))
       }
       
     }
    
  }
}

rm(outline, cell, edge, i, Outlines_0H, Outlines_1H, Outlines_2H, Outlines_3H, Outlines_4H, Outlines_6H, Outlines_all, file_, file)

# rename column names, combine file and cell info, and then keep only this unique ID
colnames(Cells_on_edge) <- c("file", "cell")
Cells_on_edge <- Cells_on_edge %>% dplyr::mutate(ID = paste0(file, "_", cell))
Cells_on_edge <- Cells_on_edge %>% select(ID)

```

IDs of cells that are touching coverslip edges (from all timepoints) are in dataframe Cells_on_edge

```{r make a list of good cells that remain after filtering}

# how many cells are with nucleus, not on edge, and not in division?

Cells_with_nucleus_not_dividing <- as.data.frame(setdiff(Cells_with_nucleus$ID, Dividing_cells$ID))
colnames(Cells_with_nucleus_not_dividing) <- "ID"

Cells_with_nucleus_not_dividing_not_edge <- as.data.frame(setdiff(Cells_with_nucleus_not_dividing$ID, Cells_on_edge$ID))
colnames(Cells_with_nucleus_not_dividing_not_edge) <- "ID"
Good_cells <- Cells_with_nucleus_not_dividing_not_edge

rm(Cells_with_nucleus_not_dividing, Cells_with_nucleus_not_dividing_not_edge)



# from these "good" cells, how many are from each timepoint? let's see e.g. at 6h

sum(grepl("_6H", Good_cells$ID))

# save good cells list

write.csv(Good_cells, "Filtered_list_of_cells.txt")

```

We start with 11807 cell before any filtering

11030 cells with nucleus
10760 cells with nucleus and not in division

There are 4 cells that are in division, but area of nucleus == 0:
"D1_2HCD3CD28_01_Cell_CP_67"  "D2_2HCD3CD28_05_Cell_CP_95"  "D3_2HCD3CD28_06_Cell_CP_80"  "D3_2HCD3CD28_08_Cell_CP_156"
These nuclei were sticking out of the cell bounds, so an error of this type was raised by fish quant:
"Nucleus could not be assigned to any cell. Must be ENTIRELY within the cell."

9432 cells with nucleus, not in division, and not on edge
There are 188 cells that are on edge, but already excluded before due to no nucleus or in-division status

So overall we lost 11807-9423 = 2384 cells with these filters (20.2% of the initial number).

From those 9423 cells that are OK after all filters, per timepoint we have:
0H - 1031
1H - 1500
2H - 1830
3H - 1613
4H - 950
6H - 2499


Next step is extracting coordinates of detected spots. While doing it, we'll only look at the cells that are in our list of "good_cells".

```{r extract spots coordinates from cells that remain after filtering}

# combine all spots files
Spots_all <- c(Spots_0H, Spots_1H, Spots_2H, Spots_3H, Spots_4H, Spots_6H) 

cytokine = ""
Spots_good = data.frame()

# go one by one logfile
for (file_ in Spots_all) {
  
  # which cytokine are the spots in this files describing? extract info from file_ name
  cytokine <- str_split(str_split(file_, "/")[[1]][9], "_")[[1]][1]
  
  # open logfile 
  spots = read.csv(file_, sep = '\t', header = FALSE, skip=13)
  colnames(spots) <- spots[1,]
  spots <- spots[-1,]
  
  # edit the table - add cytokine column and assemble ID that can be matched with Good_cells 
  spots$cytokine <- cytokine
  spots <- spots %>% dplyr::mutate(ID = substr(spots$File, 1, 15))
  spots <- spots %>% dplyr::mutate(ID = paste0(ID, "_", Cell))
  
  # keep only the relevant columns - ID, cytokine, coordinates of spots; turn coordinates to numeric cause they're characters
  spots <- spots %>% select(ID, cytokine, Z_det, Y_det, X_det)
  #sapply(spots, class) # with this you can see which class each column is
  spots$Z_det <- as.integer(spots$Z_det)
  spots$Y_det <- as.integer(spots$Y_det)
  spots$X_det <- as.integer(spots$X_det)
  
  # keep only those cells (rows) that are among Good_cells (so filtered ones)
  spots <- merge(spots, Good_cells, by = "ID", all = F)
  
  # merge all data into 1 table
  Spots_good <- rbind(Spots_good, spots)

}

rm(spots, cytokine, file_, Spots_0H, Spots_1H, Spots_2H, Spots_3H, Spots_4H, Spots_6H, Spots_all) 

write.csv(Spots_good, "Filtered_list_of_spots_coordinates.csv")

```

Spots_good now has all spots (TNF and IFN) that are within the filtered cells.

Overall across all timepoints we have:
73929 IFN spots in 6580 cells
60165 TNF spots in 5737 cells


```{r extract spots coordinates from cells that are in division}

# combine all spots files
Spots_all <- c(Spots_0H, Spots_1H, Spots_2H, Spots_3H, Spots_4H, Spots_6H) 

cytokine = ""
Spots_good = data.frame()

# go one by one logfile
for (file_ in Spots_all) {
  
  # which cytokine are the spots in this files describing? extract info from file_ name
  cytokine <- str_split(str_split(file_, "/")[[1]][9], "_")[[1]][1]
  
  # open logfile 
  spots = read.csv(file_, sep = '\t', header = FALSE, skip=13)
  colnames(spots) <- spots[1,]
  spots <- spots[-1,]
  
  # edit the table - add cytokine column and assemble ID that can be matched with Good_cells 
  spots$cytokine <- cytokine
  spots <- spots %>% dplyr::mutate(ID = substr(spots$File, 1, 15))
  spots <- spots %>% dplyr::mutate(ID = paste0(ID, "_", Cell))
  
  # keep only the relevant columns - ID, cytokine, coordinates of spots; turn coordinates to numeric cause they're characters
  spots <- spots %>% select(ID, cytokine, Z_det, Y_det, X_det)
  #sapply(spots, class) # with this you can see which class each column is
  spots$Z_det <- as.integer(spots$Z_det)
  spots$Y_det <- as.integer(spots$Y_det)
  spots$X_det <- as.integer(spots$X_det)
  
  # keep only those cells (rows) that are among dividing cells
  spots <- merge(spots, Dividing_cells, by = "ID", all = F)
  
  # merge all data into 1 table
  Spots_good <- rbind(Spots_good, spots)

}

rm(spots, cytokine, file_, Spots_0H, Spots_1H, Spots_2H, Spots_3H, Spots_4H, Spots_6H, Spots_all) 

write.csv(Spots_good, "Filtered_list_of_spots_coordinates_DIVIDING_CELLS.csv")

length(unique(Spots_good$ID)) # 236
table(Spots_good$cytokine)

rm(Spots_good)

```

4859 spots (2624 IFNG and 2235 TNF) in 236 dividing cells (from a total of 274 dividing cells that we filtered out).
With this filtered list, we can next determine their subcellular localization by combining this data with Cellpose output.
